Thermodynamic response functions in a cell fluid model

### 2024-08-12 main_morse_response_functions.tex
rewrite compressibility in terms of second pressure derivative with respect to chemical potential

	The density of the many-particle system is the ratio of the average number of particles $\left\langle N \right\rangle$  to the total volume of the system $V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to the product $\beta \mu$. Using the general relation for the equation of state $PV\beta = \ln \Xi $ the density reads
	\begin{equation}
	\eta = \frac{<N>}{v N_v} = \frac{1}{v} \frac{\partial}{\partial\beta\mu} Pv\beta = \frac{\bar n}{v},
	\end{equation} 
	where	
	\begin{equation}\label{eq:density}
	\bar n = n_g - \bar{M} + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{equation}\label{eq:crit_dens}
	n_g = g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2}.
	\end{equation}
	Combination of the equations \eqref{eq:density} and \eqref{eq:ro_M} reveals the explicit expression for the effective chemical potential $M$ as a function of the density and the temperature
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,<N>}
	\end{equation}

@@

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,<N>}
	\end{equation}

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,<N>}
	\end{equation}

>>>>

	The density of the many-particle system is the ratio of the average number of particles $\left\langle N \right\rangle$  to the total volume of the system $V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to the product $\beta \mu$. Using the general relation for the equation of state $PV\beta = \ln \Xi $ the density reads
	\begin{equation}
	\eta = \frac{\langle N\rangle}{v N_v} = \frac{1}{v} \frac{\partial}{\partial\beta\mu} Pv\beta = \frac{\bar n}{v},
	\end{equation} 
	where	
	\begin{equation}\label{eq:density}
	\bar n = n_g - \bar{M} + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{equation}\label{eq:crit_dens}
	n_g = g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2}.
	\end{equation}
	Combination of the equations \eqref{eq:density} and \eqref{eq:ro_M} reveals the explicit expression for the effective chemical potential $M$ as a function of the density and the temperature
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
	\end{equation}
	Let us perform some transformations to rewrite the definition for $K_T$ into a form that is more suitable for the equation of state~\eqref{eq:eosMT}.
	\begin{eqnarray*}
		K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T,\langle N\rangle}}
		{\left(\partial P / \partial \mu\right)_{T,\langle N\rangle}},
	\end{eqnarray*}
	where $\rho = \langle N \rangle / V$ is the particle number density.
	Applying the Gibbs--Duhem equation at $T = const$, one has
	\begin{equation*}
		{\rm d} P = \rho {\rm d} \mu,
	\end{equation*}
	or
	\begin{equation}
		\label{eq:rho_dpdm}
		\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
	\end{equation}
	Substituting this into the last expression for $K_T$, one gets
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T,\langle N\rangle}
	\end{equation}
	Finally, from~\eqref{eq:rho_dpdm} it follows that
	\begin{equation}
		\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
	\end{equation}
	and ultimately we arrive at the very useful expression for the isothermal compressibility
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T,\langle N\rangle}
	\end{equation}

@@

	\subsection{Thermal pressure coefficient}
	The thermal pressure coefficient is defined by
	\begin{equation}
		\label{def:therm_pres_coef}
		\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}
	\end{equation}

@@

	\subsection{Thermal expansion coefficient}
	The thermal expansion coefficient is defined by
	\begin{equation}
		\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}
	\end{equation}
